Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information

### 2023-01-09 write-up/covid-supplement-writeup.tex
fixing links

Table~\ref{tab:indiana} summarizes the statewide random molecular tests conducted in Indiana between April 25--29, 2020 as well as an additional non-random sample of 898 individuals during May 2--3, 2020.

\begin{table}[!th]
 \begin{tabular}{c | c | r r r | r | r r }
 & & \multicolumn{3}{c}{Total Tests (\%)} & IN & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
 \cline{3-5} \cline{7-8}
 & & NonRandom & Random & CTIS & Census & NonRandom & Random \\ \hline
 \multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 54.0 & 50.7 & 21.7 (11.2) & 1.4 \\
 & Male & 41.8 & 45.0  & 46.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
 \multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0  & 36.2 & 52.7
 & 29.7 (15.0) & 1.7 \\
 & $40-59$ & 41.1 & 36.0  & 34.3 & 25.2 & 24.9 (12.5) & 2.1 \\
 & $\geq 60$ & 19.5 & 36.0  & 29.5 & 22.1 & 6.7 (3.4) & 0.9 \\ \hline
 \multirow{2}{*}{Race} & White & 23.1 & 92.0  & - & 86.9 & 19.5 (9.6) & 1.5 \\
 & Nonwhite & 76.9 & 8.0  & - & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
 % Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
 % Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
 \multirow{2}{*}{Fever} & Yes & 17.0 & 1.8  & 1.0 & - & 66.4 (32.1) & 4.5 \\
 & No & 83.0 & 98.2  & 99.0 & - & 15.6 (7.5) & 1.3 \\ \hline
 Household & Yes & 10.8 & 1.4  & 1.8 & - & 46.1 (22.4) & 29.4 \\
 $+$ Case & No & 89.2 & 98.6  & 98.2 & - & 21.6 (10.4) & 1.3 \\ \hline
 Prior $+$ & Yes & 6.1 & 1.4  & - & - & 39.2 (20.2) & 24.4 \\
 Test & No & 93.9 & 98.6  & - & - & 21.6 (11.1) & 1.3 \\ \hline
 \end{tabular}
 \caption{Estimated total tests ($\%$) and point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\citep{Yiannoutsos2021,doi:10.1073/pnas.2111454118}. NonRandom positive test rates in parentheses are adjusted rates to match the statewide rate of $11.7\%$ on August 30th.}
 \label{tab:indiana}
 \end{table}

>>>>

Table~\ref{app:indianatable} summarizes the statewide random molecular tests conducted in Indiana between April 25--29, 2020 as well as an additional non-random sample of 898 individuals during May 2--3, 2020.

\begin{table}[!th]
 \begin{tabular}{c | c | r r r | r | r r }
 & & \multicolumn{3}{c}{Total Tests (\%)} & IN & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
 \cline{3-5} \cline{7-8}
 & & NonRandom & Random & CTIS & Census & NonRandom & Random \\ \hline
 \multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 54.0 & 50.7 & 21.7 (11.2) & 1.4 \\
 & Male & 41.8 & 45.0  & 46.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
 \multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0  & 36.2 & 52.7
 & 29.7 (15.0) & 1.7 \\
 & $40-59$ & 41.1 & 36.0  & 34.3 & 25.2 & 24.9 (12.5) & 2.1 \\
 & $\geq 60$ & 19.5 & 36.0  & 29.5 & 22.1 & 6.7 (3.4) & 0.9 \\ \hline
 \multirow{2}{*}{Race} & White & 23.1 & 92.0  & - & 86.9 & 19.5 (9.6) & 1.5 \\
 & Nonwhite & 76.9 & 8.0  & - & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
 % Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
 % Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
 \multirow{2}{*}{Fever} & Yes & 17.0 & 1.8  & 1.0 & - & 66.4 (32.1) & 4.5 \\
 & No & 83.0 & 98.2  & 99.0 & - & 15.6 (7.5) & 1.3 \\ \hline
 Household & Yes & 10.8 & 1.4  & 1.8 & - & 46.1 (22.4) & 29.4 \\
 $+$ Case & No & 89.2 & 98.6  & 98.2 & - & 21.6 (10.4) & 1.3 \\ \hline
 Prior $+$ & Yes & 6.1 & 1.4  & - & - & 39.2 (20.2) & 24.4 \\
 Test & No & 93.9 & 98.6  & - & - & 21.6 (11.1) & 1.3 \\ \hline
 \end{tabular}
 \caption{Estimated total tests ($\%$) and point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\citep{Yiannoutsos2021,doi:10.1073/pnas.2111454118}. NonRandom positive test rates in parentheses are adjusted rates to match the statewide rate of $11.7\%$ on August 30th.}
 \label{app:indianatable}
 \end{table}

### 2023-01-09 write-up/covid-writeup.tex
fixing links

 Due to testing restrictions and other potential selection biases, publicly reported COVID-19 case count data may not be sufficient to understand active infection rates or the disease trajectory.  Here, we discuss two random samples that provide auxiliary information and may help address selection bias.

@@

 Between April 25--29, 2020, Indiana conducted statewide random molecular testing of persons ages $\geq 12$ years to assess prevalence of active infection to SARS-CoV-2~\citep{Yiannoutsos2021}. A stratified random sampling design was conducted using Indiana’s 10 public health preparedness districts as sampling strata. 15,495 participants were contacted resulting in a final sample size of 3,658. Demographic data was collected (e.g., summary statistics on age, sex, and race) as well as data on whether they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2. During May 2--3, 2020, an additional non-random sample of 898 individuals was also collected.  Table~\ref{tab:indiana} in Appendix~\ref{app:in_add_details} summarizes the data.
 % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.

@@

 We start by using Table~\ref{tab:indiana} to compute IPW weights for end of April, early May.  The random sample is $n=3,658$. IPW weights are computed per strata using Indiana Census and random survey data, and are allowed to depend on Gender, Age, Race, Fever, Positive Case in Household, and Prior Positive Test. The weight for the strata defined as Male, 40--59, White, has a Fever, no household cases, and no prior tests, for example, is proportional to:
 $$
 \frac{\left( 0.493 \times 0.252 \times 0.869 \times 0.018 \times 0.986 \times 0.986 \right)}{\left(0.418 \times 0.411 \times 0.231 \times 0.17 \times 0.892 \times 0.939 \right)} \approx 0.334
 $$
 Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7\%$, a decrease of four percentage points from the observed prevalence of $11.7\%$. We also make use of CTIS sample over that window of time to construct a second IPW estimate conditional on demographics and fever but ignoring the household and prior testing information, which under no measurement-error is $6.5$\%.

@@

 There is nothing routine about COVID-19, including the corresponding statistical questions.  The goal of this paper was to point out questionable statistical routines.  Precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty. Extensions of recent statistical error decompositions~\citep{Meng2018} demonstrate how selection bias leads data analysts to feel certain about incorrect conclusions.  As case-count data is routinely used for public health policy making, we presented an inverse-probability weighting method and a doubly-robust estimation method that leverage auxiliary information collected through random samples to overcome these issues. We end with a brief discussion of important related topics.

@@

% \textcolor{red}{
 While we emphasize the importance of random sampling, we do not view it as a panacea. With access to \emph{only} a non-probabilistic sample, we cannot address self-selection bias.  Probability samples provide necessary auxiliary information to do so.  Data analysts can then focus on statistical issues related to the random samples such as how to handle non-response bias.  Covariates that the scientific team think are correlated with non-response (e.g., political affiliation, rural/urban location) should be collected as part of the random sampling protocol and corrected for in the data analysis. On the other hand, collecting covariate information that correlates with testing propensity (e.g., symptom status) in the non-probabilistic sample is insufficient.
% }

@@

 A common argument is that the SEIR model can be extended to account for selection bias and measurement error directly; therefore, there is no need for auxiliary random sampling.  Without strong assumptions on the selection mechanism, however, the estimates are often not identifiable.  When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\citep{CoxHink74}. Absent some type of random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.  \cite{Ironse2103272118}, for example, use the Indiana seroprevalence random survey and to anchor their analysis; however, infrequent random surveys imply time-varying aspects such as the IFR may not be resolved by use of a single anchor.

@@

 The proposed methodology relied on access to two critical datasets: (1) COVID-19 testing, case, hospitalization, and death data by demographic strata and (2) CTIS data.  Public access to (1) required submission of data requests to the state of Indiana.  The author's request was one of only five data requests to be approved by Indiana for public release.  Most states do not release such granular data, making selection bias adjustments difficult.  Coordinated, systematic data collection and reporting is critical.
 Lack of covariate information on individuals seeking COVID-19 tests is unacceptable.  Random samples over time to supplement this data with auxiliary information is desperately needed.  We acknowledge that this proposal will lead to data privacy concerns that will need to be addressed.

@@

 A valid criticism of the proposed approach is that recording covariates on every tested individual is time-consuming and costly.  We argue that the list of relevant factors is of reasonable length.  For example, symptom status and COVID-19 contact are clearly relevant factors in testing selection.  Future work may consider how to include/exclude factors over time that are not significant to limit reporting burden.  Moreover, equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample, which is often not feasible for state/local governments where rapid sample collection is prioritized.  One can easily extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information.
 This would balance goals of rapid testing and auxiliary data collection.

>>>>

 Due to testing restrictions and other potential selection biases, publicly reported COVID-19 case count data may not be sufficient to understand active infection rates or the disease trajectory.  Here, we discuss two samples that provide auxiliary information and may help address selection bias.

@@

 Between April 25--29, 2020, Indiana conducted statewide random molecular testing of persons ages $\geq 12$ years to assess prevalence of active infection to SARS-CoV-2~\citep{Yiannoutsos2021}. A stratified random sampling design was conducted using Indiana’s 10 public health preparedness districts as sampling strata. 15,495 participants were contacted resulting in a final sample size of 3,658. Demographic data was collected (e.g., summary statistics on age, sex, and race) as well as data on whether they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2. During May 2--3, 2020, an additional non-random sample of 898 individuals was also collected.  Table 1 in Appendix~\ref{app:in_add_details} summarizes the data.
 % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.

@@

 We start by using Table~\ref{app:indianatable} to compute IPW weights for end of April, early May.  The random sample is $n=3,658$. IPW weights are computed per strata using Indiana Census and random survey data, and are allowed to depend on Gender, Age, Race, Fever, Positive Case in Household, and Prior Positive Test. The weight for the strata defined as Male, 40--59, White, has a Fever, no household cases, and no prior tests, for example, is proportional to:
 $$
 \frac{\left( 0.493 \times 0.252 \times 0.869 \times 0.018 \times 0.986 \times 0.986 \right)}{\left(0.418 \times 0.411 \times 0.231 \times 0.17 \times 0.892 \times 0.939 \right)} \approx 0.334
 $$
 Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7\%$, a decrease of four percentage points from the observed prevalence of $11.7\%$. We also make use of CTIS sample over that window of time to construct a second IPW estimate conditional on demographics and fever but ignoring the household and prior testing information, which under no measurement-error is $6.5$\%.

@@

 There is nothing routine about COVID-19, including the corresponding statistical questions.  The goal of this paper was to point out questionable statistical routines.  Precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty. 
%Extensions of recent statistical error decompositions~\citep{Meng2018} demonstrate how selection bias leads data analysts to feel certain about incorrect conclusions.  As case-count data is routinely used for public health policy making, we presented an inverse-probability weighting method and a doubly-robust estimation method that leverage auxiliary information collected through random samples to overcome these issues. 
We end with a brief discussion of important related topics.

@@

% \textcolor{red}{
% While we emphasize the importance of random sampling, we do not view it as a panacea. With access to \emph{only} a non-probabilistic sample, we cannot address self-selection bias.  Probability samples provide necessary auxiliary information to do so.  Data analysts can then focus on statistical issues related to the random samples such as how to handle non-response bias.  Covariates that the scientific team think are correlated with non-response (e.g., political affiliation, rural/urban location) should be collected as part of the random sampling protocol and corrected for in the data analysis. On the other hand, collecting covariate information that correlates with testing propensity (e.g., symptom status) in the non-probabilistic sample is insufficient.
% }

@@

 A common argument is that the SEIR model can be extended to account for selection bias and measurement error directly; therefore, there is no need for auxiliary random sampling.  Without strong assumptions on the selection mechanism, however, the estimates are often not identifiable.  When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\citep{CoxHink74}. Absent some type of random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.  
%\cite{Ironse2103272118}, for example, use the Indiana seroprevalence random survey and to anchor their analysis; however, infrequent random surveys imply time-varying aspects such as the IFR may not be resolved by use of a single anchor.

@@

 The proposed methodology relied on access to two critical datasets: (1) COVID-19 testing, case, hospitalization, and death data by demographic strata and (2) CTIS data.  Public access to (1) required submission of data requests to the state of Indiana.  The author's request was one of only five data requests to be approved by Indiana for public release.  Most states do not release such granular data, making selection bias adjustments difficult.  Coordinated, systematic data collection and reporting is critical.
 Lack of covariate information on individuals seeking COVID-19 tests is unacceptable.  Random samples over time to supplement this data with auxiliary information is desperately needed.  We acknowledge that this proposal will lead to data privacy concerns as well as adjustments to the proposed method to account for the time-consuming and costly nature of data collection will both need to be addressed.

@@

% A valid criticism of the proposed approach is that recording covariates on every tested individual is time-consuming and costly.  We argue that the list of relevant factors is of reasonable length.  For example, symptom status and COVID-19 contact are clearly relevant factors in testing selection.  Future work may consider how to include/exclude factors over time that are not significant to limit reporting burden.  Moreover, equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample, which is often not feasible for state/local governments where rapid sample collection is prioritized.  One can easily extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information.
% This would balance goals of rapid testing and auxiliary data collection.
